Thesis Proposal

### 2022-03-23 proposal/proposal.tex
apply franck feedback on proposal

\title{Thesis Proposal}
\author{Hossein Naderi \\supervised by Eric Ruppert}
\maketitle

@@

\paragraph{}
Shared data structures have become an essential field in distributed algorithms research.
We are reaching the physical limits of how many transistors we can place on a CPU core. The industry solution to provide more computational power is to increase the number of cores of the CPU. This is why distributed algorithms have become important. It is not hard to see why multiple processes cannot update sequential data structures designed for one process. For example, consider two processes trying to insert some values into a sequential linked list simultaneously. Processes $p,q$ read the same tail node, $p$ changes the next pointer of the tail node to its new node and after that $q$ does the same. In this run, $p$'s update is overwritten. One solution is to use locks; whenever a process wants to do an update or query on a data structure, the process locks it, and others cannot use it until the lock is released. Using locks has some disadvantages; for example, one process might be slow, and holding a lock for a long time prevents other processes from progressing. Moreover, locks do not allow complete parallelism since only the one process holding the lock can make progress. Our approach is to create a lock-free queue data structure that developers can safely use without the disadvantages of locks. Also we are going to resolve a common problem in previous shared queues which makes their worst case slow.

@@

The standard safety condition is called \textit{linearizability}, which ensures that for any concurrent execution on a linearizable object, each operation should appear to take effect instantaneously at some moment between its invocation and response. Figure \ref{fig::goodexample} is an example of execution on a linearizable  queue that is initially empty. The arrow shows time, and each rectangle shows the time between the invocation and the termination of an operation. Since \texttt{Enqueue(A)} and \texttt{Enqueue(B)} are concurrent, \texttt{Enqueue(B)} may or may not take effect before \texttt{Enqueue(A)}. The execution in Figure \ref{fig::badexample} is not linearizable since \texttt{A} has been enqueued before \texttt{B}, so it has to be dequeued first.

@@

\begin{figure}[t]
  \center\includegraphics[scale=0.5]{pics/bad}
  \caption{\label{fig::badexample}An example of an execution that is not linearizable. Since \texttt{Enqueue(A)} has completed before \texttt{Enqueue(B)} is invoked the \texttt{Dequeue()} should return \texttt{A}.}
\end{figure}

There are various progress properties; the strongest is wait-freedom, and the more common is lock-freedom. An algorithm is \textit{wait-free} if each operation terminates after a finite number of its own steps. We call an algorithm \textit{lock-free} if, after a sufficient number of steps, one operation terminates. A wait-free algorithm is also lock-free but not vice versa, in an infinite run of a lock-free algorithm there might be an operation that makes progress every often but never terminates.
%From now on, we are going to focus on wait freedom.

@@

Moir, Nussbaum, and Shalev~\cite{DBLP:conf/spaa/MoirNSS05} presented a more sophisticated queue by using the elimination technique. The elimination mechanism has the dual purpose of allowing operations to complete in parallel and reducing contention for the queue. An Elimination Queue consists of an MS-queue augmented with an elimination array. Elimination works by allowing opposing pairs of concurrent operations such as an enqueue and a dequeue to exchange values when the queue is empty or when concurrent operations can be linearized to empty the queue. Their algorithm makes it possible for long-running operations to eliminate an opposing operation. The empirical evaluation showed the throughput of their work is better than MS-queue, but the worst case is still the same; in case there are $p$ concurrent enqueues, their algorithm is not better than MS-queue.

@@

\begin{figure}[t]
  \center\includegraphics[scale=0.3]{pics/baskets}
  \caption{Baskets queue idea. There is a mechanism in the algorithm that orders the operations in a basket by the order of which \texttt{CAS}es first successfully.}
\end{figure}

@@

Gidenstam, Sundell and Tsigas~\cite{DBLP:conf/opodis/GidenstamST10} introduced a new algorithm using a linked list of arrays. Global head and tail pointers point to arrays containing the first and last element in the queue and the elements when they are written by \texttt{CAS} operations and dequeued. Their data structure is lock-free. An enqueue or a dequeue searches for the head array or tail array to find the first unmarked element or last written element (see~Figure~\ref{fig::sundell}). Still, if the head array is empty and $p$ processes try to enqueue simultaneously, the step complexity remains $\Omega(p)$.

@@

Kogan and Petrank~\cite{DBLP:conf/ppopp/KoganP11} introduced wait-free queues based on MS-queue and use Herlihy's helping technique to achieve wait-freedom. Their step complexity is $\Omega(p)$ because of the helping mechanism.

@@

The goal here is to ensure in each propagate step the new operations are propagated up to the parent in $O(\log p)$ steps (see~Figure~\ref{fig::propagstep}). Then, a dequeue operation uses the linearization ordering to compute its answer.

In each propagate step, our algorithm uses a subroutine \textsc{Refresh}($n$) that aggregates new operations from node $n$'s children (that have not already been propagated to $n$) and tries to append them into $n$ using a\texttt{CAS} operation. The general idea is that if we call \textsc{Refresh}($n$) twice, the operations in \texttt{n}'s children before the first \textsc{Refresh}($n$) are guaranteed to be in $n$.
%Since there is a successful instance of \textsc{Refresh}($n$) after the first \textsc{Refresh}($n$), which has appended the talked operations.
Instead of storing operations explicitly in the nodes, we only keep track of the number of them. This allows us to \texttt{CAS} fixed-size objects in each \textsc{Refresh}($n$). To do that, we introduce blocks that only contain the number of operations from the left and the right child in a \textsc{Refresh}() procedure and only propagate the block of the new operations.

@@

We also implement methods \texttt{Get(i)}, \texttt{Index(op)} to get the $i$th propagated operation and compute the rank of a propagated operation in the linearization. \texttt{Get(i)} finds the block containing the $i$th operation in the root and then finds its sub-block recursively to reach a leaf. \texttt{Index()} is  similar but more complicated, finding super-blocks from a leaf to the root. The main challenge in each level of \texttt{Get(i)} and \texttt{Index(op)} is that it should take polylogarithmic steps with respect to $p$. After appending operations to the root, processes can find out information about the linearization ordering using \texttt{Get(i)} and \texttt{Index(op)}. Now we can implement Enqueue and Dequeue using our block tree. An \texttt{Enqueue(e)} appends an operation with input argument \texttt{e} in the block tree. To do a \texttt{Dequeue()}, first process $p$ first appends a \texttt{DEQ} operation to the tree. Then $p$ finds the rank of the \texttt{DEQ} using \texttt{Index()}, the rank of the \texttt{DEQ} and the information stored in the root about the queue $p$ computes the rank of  the \texttt{ENQ} having the answer of the \texttt{DEQ}. Finally $p$ returns the argument of that \texttt{ENQ} using \texttt{Get(i)}.

>>>>

\title{Thesis Proposal}
\author{Hossein Naderibeni \\supervised by Eric Ruppert}
\maketitle

@@

\paragraph{}
Shared data structures have become an essential field in distributed algorithms research.
We are reaching the physical limits of how many transistors we can place on a CPU core. The industry solution to provide more computational power is to increase the number of cores of the CPU. This is why distributed algorithms have become important. It is not hard to see why multiple processes cannot update sequential data structures designed for one process. For example, consider two processes trying to insert some values into a sequential linked list simultaneously. Processes $p,q$ read the same tail node, $p$ changes the next pointer of the tail node to its new node and after that $q$ does the same. In this run, $p$'s update is overwritten. One solution is to use locks; whenever a process wants to do an update or query on a data structure, the process locks it, and others cannot use it until the lock is released. Using locks has some disadvantages; for example, one process might be slow, and holding a lock for a long time prevents other processes from progressing. Moreover, locks do not allow complete parallelism since only the one process holding the lock can make progress. Our approach is to create a lock-free queue data structure that developers can safely use without the disadvantages of locks.
%We are going to resolve a common problem in previous shared queues which makes their worst case slow.

@@

The standard safety condition is called \textit{linearizability}, which ensures that for any concurrent execution on a linearizable object, each operation should appear to take effect instantaneously at some moment between its invocation and response. Figure \ref{fig::goodexample} is an example of an execution on a linearizable  queue that is initially empty. The arrow shows time, and each rectangle shows the time between the invocation and the termination of an operation. Since \texttt{Enqueue(A)} and \texttt{Enqueue(B)} are concurrent, \texttt{Enqueue(B)} may or may not take effect before \texttt{Enqueue(A)}. The execution in Figure \ref{fig::badexample} is not linearizable since \texttt{A} has been enqueued before \texttt{B}, so it has to be dequeued first.

@@

\begin{figure}[t]
  \center\includegraphics[scale=0.5]{pics/bad}
  \caption{\label{fig::badexample}An example of an execution that is not linearizable. Since \texttt{Enqueue(A)} has completed before \texttt{Enqueue(B)} is invoked the \texttt{Dequeue()} should return \texttt{A} or nothing.}
\end{figure}

There are various progress properties; the strongest is wait-freedom, and the more common is lock-freedom. An algorithm is \textit{wait-free} if each operation terminates after a finite number of its own steps. We call an algorithm \textit{lock-free} if, after a sufficient number of steps, one operation terminates. A wait-free algorithm is also lock-free but not vice versa, in an infinite run of a lock-free algorithm there might be an operation that makes progress every so often but never terminates.
%From now on, we are going to focus on wait freedom.

@@

Moir, Nussbaum, and Shalev~\cite{DBLP:conf/spaa/MoirNSS05} presented a more sophisticated queue by using the elimination technique. The elimination mechanism has the dual purpose of allowing operations to complete in parallel and reducing contention for the queue. An Elimination Queue consists of an MS-queue augmented with an elimination array. Elimination works by allowing opposing pairs of concurrent operations such as an enqueue and a dequeue to exchange values when the queue is empty or when concurrent operations can be linearized to empty the queue. Their algorithm makes it possible for long-running operations to eliminate an opposing operation. The empirical evaluation showed the throughput of their work is better than the MS-queue, but the worst case is still the same; in case there are $p$ concurrent enqueues, their algorithm is not better than MS-queue.

@@

\begin{figure}[t]
  \center\includegraphics[scale=0.3]{pics/baskets}
  \caption{Baskets queue idea. There is a time that all operations in a basket were running concurrently, but only one has succeeded to do \texttt{CAS}. To order the operations in a basket, the mechanism in the algorithm for processes is to \texttt{CAS} again. The successful process will be the next one in the basket and so on.}
\end{figure}

@@

Gidenstam, Sundell, and Tsigas~\cite{DBLP:conf/opodis/GidenstamST10} introduced a new algorithm using a linked list of arrays. Global head and tail pointers point to arrays containing the first and last elements in the queue. Global pointers are up to date, but head and tail pointers may be behind in time. An enqueue or a dequeue searches in the head array or tail array to find the first unmarked element or last written element (see~Figure~\ref{fig::sundell}). Their data structure is lock-free. Still, if the head array is empty and $p$ processes try to enqueue simultaneously, the step complexity remains $\Omega(p)$.

@@

Kogan and Petrank~\cite{DBLP:conf/ppopp/KoganP11} introduced wait-free queues based on the MS-queue and use Herlihy's helping technique to achieve wait-freedom. Their step complexity is $\Omega(p)$ because of the helping mechanism.

@@

The goal here is to ensure that in each propagate step the new operations are propagated up to the parent in $O(\log p)$ steps (see~Figure~\ref{fig::propagstep}). Then, a dequeue operation uses the linearization ordering to compute its answer.

In each propagate step, our algorithm uses a subroutine \textsc{Refresh}($n$) that aggregates new operations from node $n$'s children (that have not already been propagated to $n$) and tries to append them into $n$ using a \texttt{CAS} operation. The general idea is that if we call \textsc{Refresh}($n$) twice, the operations in $n$'s children before the first \textsc{Refresh}($n$) are guaranteed to be in $n$.
%Since there is a successful instance of \textsc{Refresh}($n$) after the first \textsc{Refresh}($n$), which has appended the talked operations.
Instead of storing operations explicitly in the nodes, we only keep track of the number of them. This allows us to \texttt{CAS} fixed-size objects in each \textsc{Refresh}($n$). To do that, we introduce blocks that only contain the number of operations from the left and the right child in a \texttt{Refresh}() procedure and only propagate the block of the new operations.

@@

We also implement methods \texttt{Get(i)}, \texttt{Index(op)} to get the $i$th propagated operation and compute the rank of a propagated operation in the linearization. \texttt{Get(i)} finds the block containing the $i$th operation in the root and then finds its sub-block recursively to reach a leaf. \texttt{Index()} is  similar but more complicated, finding super-blocks from a leaf to the root. The main challenge in each level of \texttt{Get(i)} and \texttt{Index(op)} is that it should take polylogarithmic steps with respect to $p$. After appending operations to the root, processes can find out information about the linearization ordering using \texttt{Get(i)} and \texttt{Index(op)}. Now we can implement Enqueue and Dequeue using our block tree. An \texttt{Enqueue(e)} appends an operation with input argument \texttt{e} in the block tree. To do a \texttt{Dequeue()}, process $p$ first appends a \texttt{DEQ} operation to the tree. Then $p$ finds the rank of the \texttt{DEQ} using \texttt{Index()}, the rank of the \texttt{DEQ} and the information stored in the root about the queue $p$ computes the rank of  the \texttt{ENQ} having the answer of the \texttt{DEQ}. Finally $p$ returns the argument of that \texttt{ENQ} using \texttt{Get(i)}.
